Type-Safe Modular Parsing

### 2017-01-09 paper/sections/Discussion.tex
Related work ready for review.

\subsection{Alternative Techniques}
\huang{to rewrite}
In the previous sections we have talked about Packrat Parsing, open recursion and Object Algebras. These components fuse together in our library implementation, for the initial goal, which is to build extensible parsers. We want to argue here that those aspects are totally orthogonal, and hence can have their alternatives. For example, Object Algebras are helpful to build extensible data structures, whereas in functional programming, there has been a large amount of related work on extensible datatypes, including DTC and MRM for the Haskell language. And a better
parsing library would potentially introduce a lot of fancy features and parser combinators, run with high efficiency, or even support indirect left-recursion.
\haoyuan{Open recursion could also have alternatives?} For now, our implementation integrates the existing ones, and has turned out to be practical in various applications.

>>>>

\subsection{Alternative Techniques}
In the previous sections we have talked about Packrat Parsing, inheritance and Object Algebras. These components fuse together in our library implementation, for building extensible parsers. We want to argue here that those aspects are totally orthogonal, and hence can have their alternatives.

For example, we have conducted some previous experiments of modular paring on the Haskell language, based on open recursion, MRM [] and Parsec. \haoyuan{I mention our previous work here. } The delegation-based open recursion can be an alternative for inheritance. It exposes recursion explicitly by a parameter for ``self-reference''.
Furthermore, there has been a lot of related work that realizes extensible data types in functional programming, including MRM using a list of functors. We used Parsec as the parsing library, but suffered from the issues already illustrated in Section~\ref{subsec:challenges}. As a workaround, we made use of state monad in Parsec to ensure that a left-recursive production rule will not be applied successively in parsing. Whereas a better
parsing library would potentially introduce a lot of fancy features and parser combinators, run with high efficiency, or even support indirect left-recursion.

### 2017-01-09 paper/sections/PackratParsing.tex
Changes on section 3, 4
Polishing section 4 and 5
Related work ready for review.

\subsection{Algorithmic Challenges of Modularity}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of them are
problematic in a modular setting.

@@

\noindent However, this results in the worst case exponential time complexity, if the library does not have related optimization.

>>>>

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of them are
problematic in a modular setting.

@@

\noindent However, this results in the worst case exponential time complexity in Parsec, because it does not have related optimization.

### 2017-01-09 paper/sections/RelatedWork.tex
Related work ready for review.

Our work integrates several components including extensible parsing,
parser combinators and extensibility techniques. There has been a
great amount of related papers on those hot topics, of which some
inspired us of this paper and encourages us for more exploration. This
section will try to lead a discussion on what difference we have made.

\paragraph{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, of which parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with their affects on the performance.
\bruno{what performance issue?}\haoyuan{Revised}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which focused little on separate
compilation\bruno{focused little or did not have it at all?}.\haoyuan{The authors have
mentioned separate compilation as a possible yet unimportant feature; they decided to discard it.}
Extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition. Although overriding existing production
rules is tricky with our approach, we could perhaps make use of
embedded domain-specific languages on top of parser combinators and
transformations for overriding, but it is anyway an orthogonal issue.
\bruno{Not sure that I understand the last sentence}
\haoyuan{???} \haoyuan{BTW what about Racket?} \bruno{Have you read
Languages as Libraries? That is probably an important reference, which
I think we should cite.}\haoyuan{what about
  metafront? it is a macro system but does it have type safety?}
\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
  syntax" proposes a system for extensible syntax, where users write
  EDSLs in their language with concrete syntax. Users can write rules
  for type-based disambiguation. But separate compilation is again not
  mentioned. Shall we mention that thesis?} \haoyuan{attribute
  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, where grammars\bruno{what
  is a paper grammar?}\haoyuan{Revised} are compiled
to the generation of parse tables\bruno{I don't understand the last
  part; please rephrase. Also please break the sentence into multple
  sentences. Generally speaking go over the text you wrote and watch
  out for long sentences. You have a tendency to write very long
  sentences. }, which are DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive\bruno{expensive in terms of what? why is it expensive?}.
\haoyuan{in terms of performance. The sentence is copied from original paper.}
Furthermore, \name supports \emph{both} separate compilation
as well as \emph{modular type-checking}. The extensibility of
parsing in our approach is further available at language composition
and lexical level. \haoyuan{I mentioned one shortcoming of our
  approach here, which is we cannot override existing production
  rules; another one is that we do not have explicit
  correspondence/relationship between abstract syntax and the parser.}

\paragraph{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
left-recursion is known as a big issue in recursive descent parsers,
in which case programmers using libraries like Parsec have to manually
eliminate left-recursion in the grammar, which is cumbersome, but also
it distorts the shape of its old grammar, restricting the modularity
of parsing to extreme extent. It is possible to avoid some of
 the limitations of parser combinators.
\bruno{I'm not sure that th following text should be here. It could
be in the paper, but somewhere else.}
For example, we have
conducted our previous experiment in Haskell, by using open recursion,
MRM~\cite{Oliveira2015} and Parsec. To support direct left-recursive
grammars, we made use of the state monad in Parsec to ensure that a
left-recursive production rule will not be applied successively in
parsing. It turned out to be successful, yet introduced complexity in
programming, hence we believe that the parsing library is obligated
for left-recursion.

Some more recent papers~\cite{Ford2002,Might2011,Frost2008}
\haoyuan{Pracitical, general parser combinators?} proposed a series of
novel parsing techniques that addressed the issue
of left-recursion. We have selected Packrat
Parsing for our prototype implementation, as the combinators are
convenient to use in Scala, which is a platform where functional
laziness and Object Algebras can perfectly coexist. Also, its support
for direct left-recursion is already helpful for designing real-world
parsers practically, though~\cite{warth2008} further demonstrated that
general left-recursion is supported from the theoretical point of
view. We have also mentioned that the components used in our library
can have alternatives, for instance, a different set of parser
combinators may support ambiguous and left-recursive grammars, or be
applicable to a different subset of context-free grammars or more, or
even lead us to another level of performance.

To the best of our knowledge, no previous work on parser combinators
deals with semantic modularity and
extensibility of parsing. The modularity of abstract syntax trees is
necessary for modular parsing that supports separate compilation and
modular type-checking. In our library, Object Algebras have been
adopted for extensible ASTs, together with the behaviors on them.

\paragraph{Extensibility} There has been alot of work on building
modular programs, in both functional and object-oriented programming .
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. It supports extensible datatypes, as well as
the operations on them by implementing instances of functors. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. As mentioned above, we have conducted our early experiment
with MRM for modular datatypes, whereas the complexity of library
implementation was comparably high. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.

In OO languages, however, things get much better with the mechanism of
inheritance. Furthermore, Object Algebras~\cite{Oliveira2012} were
proposed as a design pattern for the extensibility of both data
variants and operations in a modular type-safe way. The code is
concise and expressive using Object Algebras, building a foundation
for our new library with various dimensions of extensibility.
Moreover, in~\cite{Oliveira2012} the authors have discussed the
composition of algebras. In our parsing approach, a parser consumes an
algebra, which is delegated to return the results, during its process
of parsing. Having a set of algebras, it requires multiple parsing
with several times of invocation, which leads to redundant work.
Instead, algebras are supposed to be composed into one before the
invocation of the parser. Bahr et al. lead a similar discussion
in~\cite{Bahr2011}, where queries (or \textit{catamorphisms}) and
transformations (or \textit{homomorphisms}) are composable. They have
also mentioned the dual process of folds, namely
\textit{anamorphisms}. It is potentially related to our work, as
parsing is a representative kind of unfolds, whereas they only
discussed the composition of a cv-coalgebra and a term homomorphism,
which differs from modular parsing. \haoyuan{From Object Algebras to
  Attribute Grammars?}

Another interesting observation, as we have mentioned
in~\ref{subsec:objectalgebras}, is that transformation algebras are
frequently used in parsing, as concrete syntax may not always coincide
with its abstract syntax, for instance, the bruijn indices in lambda
calculus, as well as other syntactic sugars. In the Shy
framework~\cite{Zhang2015}, the pattern of transformations is captured
into a template for Java annotation processing, and most of
boilerplate code is automatically generated, hence users only need to
override a few interesting cases, conveniently. The idea of overriding
existing cases was also proposed in MRM. It is potentially useful to
our framework.\bruno{The discussion on Extensibility can be
  considerably shorter. We just need to point out that there are many
  techniques for extensibility (without discussing alot of details),
  but most of that work does not address the problem of
\emph{producing extensible ASTs} (rather than consuming those. Then
you can talk about Baht's work, and have the discussion on
transformations. Generally speaking, go over your text, polish it, and
trim it. You are a bit too ``chatty'' (that is you talk too much): go
straight to the point.}

>>>>

Our work integrates several components including extensible parsing,
parser combinators and extensibility techniques. There has been a
great amount of related papers on those hot topics. Some
inspired us of this paper and encourage us for more exploration. This
section will try to lead a discussion on what difference we have made.

\paragraph{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, among them parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with their affects on the performance.
\bruno{what performance issue?}\haoyuan{Revised}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which did not focus on separate
compilation.\bruno{focused little or did not have it at all?}.\haoyuan{The authors have
mentioned separate compilation as a possible yet unimportant feature; they decided to discard it.}
Racket adopts macro system for library-based language extensibility. It is powerful with attributed
ASTs for contextual information, and extensions can be integrated in a modular way. Whereas
such modularity is not flexible enough for language unification, as the syntax is only built from extensions.
Moreover, existing macros cannot be further changed after definition. \haoyuan{correct?}
On the other hand, extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition.
%\bruno{Have you read
%Languages as Libraries? That is probably an important reference, which
%I think we should cite.}
%\haoyuan{what about
%  metafront? it is a macro system but does it have type safety?}
%\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
%  syntax" proposes a system for extensible syntax, where users write
%  EDSLs in their language with concrete syntax. Users can write rules
%  for type-based disambiguation. But separate compilation is again not
%  mentioned. Shall we mention that thesis?} \haoyuan{attribute
%  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, where grammars are compiled
to the generation of parse tables. Those parse tables are expressed as DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive in terms of performance.
Furthermore, our approach supports \emph{both} separate compilation
as well as \emph{modular type-checking}. The extensibility of
parsing is further available at language composition
and lexical level.
%\haoyuan{we do not have explicit
%  correspondence/relationship between abstract syntax and the parser.}

\paragraph{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination, back-tracking and
longest match composition issues, as we discussed before.

Some more recent papers~\cite{Ford2002,Might2011,Frost2008} proposed a series of
novel parsing techniques that addressed the issue
of left-recursion. We have selected Packrat
Parsing for our prototype implementation, as the combinators are
convenient to use in Scala, which is a platform where functional
laziness and Object Algebras can perfectly coexist. Also, its support
for direct left-recursion is already helpful for designing real-world
parsers practically, though~\cite{warth2008} further demonstrated that
general left-recursion is supported theoretically.
Yet we have mentioned that the components used in our approach
can have alternatives to achieve better modular parsing.

\paragraph{Extensibility} To the best of our knowledge, none of the existing techniques
address the problem of producing extensible ASTs. Certainly, there has been a lot of work on extensible
data types, in both functional and object-oriented programming.
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.

In OO languages, however, things get much better with the mechanism of
inheritance. Furthermore, Object Algebras~\cite{Oliveira2012} were
proposed as a design pattern for the extensibility of both data
variants and operations in a modular type-safe way. The code is
concise and expressive using Object Algebras, building a foundation
for our new library with various dimensions of extensibility.
Moreover, in~\cite{Oliveira2012} the authors have discussed the
composition of algebras. In our parsing approach, a parser consumes an
algebra, which is delegated to return the results, during its process
of parsing. Having a set of algebras, it requires multiple parsing
with several times of invocation, which leads to redundant work.
Instead, algebras are supposed to be composed into one before the
invocation of the parser. Bahr et al. lead a similar discussion
in~\cite{Bahr2011}, where queries (or \textit{catamorphisms}) and
transformations (or \textit{homomorphisms}) are composable. They have
also mentioned the dual process of folds, namely
\textit{anamorphisms}. It is potentially related to our work, as
parsing is a representative kind of unfolds, whereas they only
discussed the composition of a cv-coalgebra and a term homomorphism,
which differs from modular parsing.

Another interesting observation, as we have mentioned
in~\ref{subsec:objectalgebras}, is that transformation algebras are
frequently used in parsing, as concrete syntax may not always coincide
with its abstract syntax, for instance, the bruijn indices in lambda
calculus, as well as other syntactic sugars. In the Shy
framework~\cite{Zhang2015}, the pattern of transformations is captured
into a template for Java annotation processing, and most of
boilerplate code is automatically generated, hence users only need to
override a few interesting cases, conveniently. It is potentially useful
in our case. By the way, the idea of overriding
existing cases was also proposed in MRM. Our approach allows existing parsing
code to be overridden during inheritance.
